P4IRS: An Intermediate Representation and Compiler for Parallel and Performance-Portable Particle Simulations

### 2023-11-18 manuscript.tex
Add optimization strategies
Signed-off-by: Rafael Ravedutti <rafael.r.ravedutti@fau.de>

@@

@@

>>>>

To obtain the force for a particle in a specific time-step, it is necessary to compute its partial contributions from each of the other particles present in the system.
In a naive strategy, this results in a quadractic complexity and severely affects the application performance, which limits the system sizes and time-scales of such simulations.

When force contributions become negligible at long-distance interactions, different algorithm strategies can be employed to perform computations only for particles within a specific cut-off region.
There are two main strategies which are commonly used for particle simulations in state-of-the-art packages: the Linked Cells approach and the Verlet Lists.

Linked Cells strategy sorts atoms spatially according to their position in cells (also called bins), where cells have a fixed size, and mapping particles into cells can be done straightforward by computing in which cell a particle is contained within.
After, forces can be computed for each particle by traversing its current and neighbor cells, and only adding contributions for particles within these cells.

The Linked Cells results in computing only forces for particles within a region that is discretized by their current and neighbor cells/bins, which can still results in a significant amount of redundant computations.
In order avoid even more redundant computations, the Verlet Lists strategy can be used jointly with Linked Cells.
The Verlet Lists constructs a list of neighbor particles for every particle, and this list can be created by using the Linked Cells structure, and doing particle-particle distance comparison to evaluate wether a particle should be added into the list.
At the end, the region for each particle search is defined by a sphere.
Note that, for the neighbor-lists construction to be worthwhile, it should be used more than once force calculation, hence neighbor-lists are not created at every simulation time-step, and a "skin" can be added to the cutoff radius to extend the neighbor-search and avoid missing computations from atoms that end up entering the cut-off region.

Despite avoiding computations from long-distance particles, another common optimization is to take advantage of Newton's Third Law and computing forces in only one direction.
This means that only half of particle-pairs are traversed during force computation, and forces are decreased for the neighbor atom during their computation.
Ideally, this should lead to a factor of two speedup, but this strategy has a few drawbacks: (a) parallelism is harmed since decreasing force of j-atoms requires atomic operations to avoid race conditions and (b) it is required to gather and scatter forces for j-atoms into a SIMD fashion, which results in scattered accesses and extra instructions to properly organize the SIMD vectors.
At the end, this strategy is likely to be beneficial for more arithmetic-intensive kernels, and preferably when the trade-off between parallelism and less arithmetic operation is compensated.

% Linked Cells, Verlet Lists, Cluster Pair
% Half-neighbor lists, Full-neighbor lists

@@

% Show examples (LJ, EAM)
% Linked Cells / Verlet Lists
% Can generate CPU/GPU code
% Vectors, quaternions and matrices
% Properties, feature properties and contact properties
% Contact history
% Shape partitioning

@@

% Testbed

% Single core CPU vs. MD-Bench (Runtime, MFLOPS/s, ...?)
% CPU MD weak-scaling
% Single GPU (vs. MD-Bench?)
% GPU MD weak-scaling

% DEM vs MESA-PD on CPU
%  - Linked Cells and Neighbor-lists
% DEM on a single GPU
% GPU DEM weak-scaling (or strong scaling?)
